---
title: "ANOVA In-Class Activity"
author: "Hrithik Sagar @ 2023900021"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview of Steps (from slide)

1. Check for normality  
2. Check homogeneity of variances → **Factorial / Independent ANOVA**  
   Check sphericity of variances → **Repeated Measures ANOVA**  
3. Choose the appropriate test  
4. Only if F is significant → run post-hoc test  
5. Report effect size  
6. Plot the data  

---

# Q1: Memory Scores
**Do children with neurodevelopmental disorders have lower memory scores?**  
Design: **One-Way Independent ANOVA** (different participants per group)

## Data

```{r}
regular  <- c(24, 22, 19, 22, 28, 26, 28, 24, 30, 29, 25, 20, 17, 19, 18, 26, 27, 24, 27, 27)
autism   <- c(15, 2, 1, 21, 3, 10, 9, 8, 3, 7, 6, 18, 2, 5, 2, 5, 0, 27)
epilepsy <- c(30, 15, 34, 26, 14, 28, 17, 29, 25, 11, 37, 36, 34, 22, 18, 5, 12, 10, 15)

# Combine into a data frame
score <- c(regular, autism, epilepsy)
group <- factor(rep(c("Regular", "Autism", "Epilepsy"),
                    times = c(length(regular), length(autism), length(epilepsy))))
df <- data.frame(score, group)

# Descriptive statistics per group
tapply(df$score, df$group, function(x) round(c(n=length(x), mean=mean(x), sd=sd(x)), 2))
```

## Step 1: Check Normality (Shapiro-Wilk + Q-Q Plots)

```{r fig.width=9, fig.height=3.5}
# Shapiro-Wilk test per group
for (g in levels(df$group)) {
  x <- df$score[df$group == g]
  sw <- shapiro.test(x)
  cat(g, ": W =", round(sw$statistic, 3), ", p =", round(sw$p.value, 3),
      ifelse(sw$p.value > 0.05, "-> Normal", "-> NOT Normal"), "\n")
}

# Q-Q plots
par(mfrow = c(1, 3))
for (g in levels(df$group)) {
  x <- df$score[df$group == g]
  qqnorm(x, main = paste("Q-Q:", g))
  qqline(x, col = "red")
}
par(mfrow = c(1, 1))
```

## Step 2: Check Homogeneity of Variances

```{r}
# Bartlett's test (assumes normality)
bartlett.test(score ~ group, data = df)

# Levene's test (robust to non-normality) - manual, no extra package
levene_manual <- function(y, group) {
  group_medians <- tapply(y, group, median)      # median-based for robustness
  z <- abs(y - group_medians[group])             # absolute deviations
  anova(lm(z ~ group))
}
levene_manual(df$score, df$group)
```

> If p > .05: variances are homogeneous → proceed with standard ANOVA

## Step 3: One-Way ANOVA

```{r}
model <- aov(score ~ group, data = df)
summary(model)
```

## Step 4: Effect Size (η² — Eta-squared)

```{r}
# eta-squared = SS_between / SS_total
ss      <- summary(model)[[1]][["Sum Sq"]]
eta_sq  <- ss[1] / sum(ss)
cat("Eta-squared (eta^2) =", round(eta_sq, 3), "\n")
cat("Benchmarks: small = 0.01, medium = 0.06, large = 0.14\n")
```

## Step 5: Post-Hoc Test (Tukey HSD — only if F significant)

```{r}
TukeyHSD(model)
```

## Step 6: Plot

```{r fig.width=6, fig.height=5}
boxplot(score ~ group, data = df,
        col  = c("lightblue", "lightgreen", "lightyellow"),
        main = "Memory Scores by Group",
        xlab = "Group", ylab = "Memory Score", las = 1)

# Mark group means
means <- tapply(df$score, df$group, mean)
points(1:length(means), means, pch = 18, col = "red", cex = 1.5)
legend("topright", legend = "Group mean", pch = 18, col = "red")
```

---

# Q2: Driving Ability
**What extent of sleep deprivation affects driving ability?**

```{r}
night1 <- c(15, 18, 20, 15, 12, 18, 16, 17, 14, 19, 20, 15, 16, 18, 19, 15, 17, 18, 19, 15)
night2 <- c(10, 16, 13, 11,  9, 14, 13, 14, 15, 14, 12, 13, 14, 12, 11, 15, 14, 16, 11, 14)
night3 <- c( 5,  3,  9,  6,  4,  7,  8,  2,  4,  6,  9,  5,  3,  7,  1,  8,  7,  3,  7,  3)
```

---

## Condition 1: Same participants in all 3 groups → Repeated Measures ANOVA

### Step 1: Normality

```{r}
data_list <- list("1 Night" = night1, "2 Nights" = night2, "3 Nights" = night3)
for (nm in names(data_list)) {
  sw <- shapiro.test(data_list[[nm]])
  cat(nm, ": W =", round(sw$statistic, 3), ", p =", round(sw$p.value, 3),
      ifelse(sw$p.value > 0.05, "-> Normal", "-> NOT Normal"), "\n")
}
```

### Step 2: Sphericity (Mauchly's Test)

```{r}
# mauchly.test() is built into base R
Y     <- cbind(night1, night2, night3)
idata <- data.frame(deprivation = factor(c("1_Night", "2_Nights", "3_Nights")))
fit   <- lm(Y ~ 1)
mauchly.test(fit, X = ~ 1, idata = idata)
```

> If p > .05: sphericity holds → standard RM-ANOVA F values are valid  
> If p < .05: sphericity violated → apply Greenhouse-Geisser correction (epsilon correction on df)

### Step 3: Repeated Measures ANOVA

```{r}
n     <- length(night1)
df_rm <- data.frame(
  subject     = factor(rep(1:n, 3)),
  deprivation = factor(rep(c("1_Night","2_Nights","3_Nights"), each = n),
                       levels = c("1_Night","2_Nights","3_Nights")),
  score       = c(night1, night2, night3)
)

# Error(subject/deprivation) partitions within-subject variance
rm_model <- aov(score ~ deprivation + Error(subject/deprivation), data = df_rm)
summary(rm_model)
```

### Step 4: Effect Size (Partial η²)

```{r}
rm_sum <- summary(rm_model)
ss_dep <- rm_sum$`Error: subject:deprivation`[[1]]["deprivation", "Sum Sq"]
ss_res <- rm_sum$`Error: subject:deprivation`[[1]]["Residuals",  "Sum Sq"]
eta_rm <- ss_dep / (ss_dep + ss_res)
cat("Partial Eta-squared =", round(eta_rm, 3), "\n")
```

### Step 5: Post-Hoc (Bonferroni-corrected paired t-tests)

```{r}
pairwise.t.test(df_rm$score, df_rm$deprivation,
                paired          = TRUE,
                p.adjust.method = "bonferroni")
```

### Step 6: Plot

```{r fig.width=6, fig.height=5}
means_rm <- tapply(df_rm$score, df_rm$deprivation, mean)
se_rm    <- tapply(df_rm$score, df_rm$deprivation,
                   function(x) sd(x) / sqrt(length(x)))

plot(1:3, means_rm, type = "b", pch = 19, col = "steelblue",
     ylim = c(0, max(means_rm + se_rm) + 2),
     xaxt = "n",
     main = "Driving Ability - Repeated Measures",
     xlab = "Sleep Deprivation", ylab = "Mean Score")
axis(1, at = 1:3, labels = c("1 Night","2 Nights","3 Nights"))

# Error bars (±1 SE)
arrows(1:3, means_rm - se_rm, 1:3, means_rm + se_rm,
       angle = 90, code = 3, length = 0.08, col = "steelblue")
```

---

## Condition 2: Different participants in all 3 groups → Independent One-Way ANOVA

### Step 1: Normality

```{r}
for (nm in names(data_list)) {
  sw <- shapiro.test(data_list[[nm]])
  cat(nm, ": W =", round(sw$statistic, 3), ", p =", round(sw$p.value, 3),
      ifelse(sw$p.value > 0.05, "-> Normal", "-> NOT Normal"), "\n")
}
```

### Step 2: Homogeneity of Variances

```{r}
score_ind <- c(night1, night2, night3)
cond_ind  <- factor(rep(c("1_Night","2_Nights","3_Nights"), each = 20),
                    levels = c("1_Night","2_Nights","3_Nights"))

bartlett.test(score_ind ~ cond_ind)
levene_manual(score_ind, cond_ind)
```

### Step 3: One-Way ANOVA

```{r}
model_ind <- aov(score_ind ~ cond_ind)
summary(model_ind)
```

### Step 4: Effect Size (η²)

```{r}
ss_ind  <- summary(model_ind)[[1]][["Sum Sq"]]
eta_ind <- ss_ind[1] / sum(ss_ind)
cat("Eta-squared =", round(eta_ind, 3), "\n")
```

### Step 5: Post-Hoc (Tukey HSD)

```{r}
TukeyHSD(model_ind)
```

### Step 6: Plot

```{r fig.width=6, fig.height=5}
boxplot(score_ind ~ cond_ind,
        col  = c("lightyellow", "lightblue", "lightcoral"),
        main = "Driving Ability - Independent Groups",
        xlab = "Sleep Deprivation", ylab = "Score", las = 1)

means_ind <- tapply(score_ind, cond_ind, mean)
points(1:3, means_ind, pch = 18, col = "red", cex = 1.5)
legend("topright", legend = "Group mean", pch = 18, col = "red")
```

---

# Q3: Exam Performance
**Does school type affect exam scores?**  
Design: **One-Way Independent ANOVA** (different participants per group)

## Data

```{r}
home_school <- c(89, 75, 49, 87, 84, 68, 88, 78, 77, 93, 67, 79, 69, 88, 91)
boarding    <- c(85, 78, 59, 77, 63, 88, 71, 73, 69, 80, 72, 68, 66, 59, 70)
regular_day <- c(91, 88, 84, 81, 91, 75, 69, 93, 95, 85, 87, 84, 83, 80, 77)

score_exam <- c(home_school, boarding, regular_day)
school     <- factor(rep(c("Home","Boarding","Regular Day"), each = 15),
                     levels = c("Home","Boarding","Regular Day"))
df_exam    <- data.frame(score_exam, school)

tapply(df_exam$score_exam, df_exam$school,
       function(x) round(c(n=length(x), mean=mean(x), sd=sd(x)), 2))
```

## Step 1: Normality

```{r fig.width=9, fig.height=3.5}
for (g in levels(df_exam$school)) {
  x  <- df_exam$score_exam[df_exam$school == g]
  sw <- shapiro.test(x)
  cat(g, ": W =", round(sw$statistic, 3), ", p =", round(sw$p.value, 3),
      ifelse(sw$p.value > 0.05, "-> Normal", "-> NOT Normal"), "\n")
}

par(mfrow = c(1, 3))
for (g in levels(df_exam$school)) {
  x <- df_exam$score_exam[df_exam$school == g]
  qqnorm(x, main = paste("Q-Q:", g))
  qqline(x, col = "red")
}
par(mfrow = c(1, 1))
```

## Step 2: Homogeneity of Variances

```{r}
bartlett.test(score_exam ~ school, data = df_exam)
levene_manual(df_exam$score_exam, df_exam$school)
```

## Step 3: One-Way ANOVA

```{r}
model_exam <- aov(score_exam ~ school, data = df_exam)
summary(model_exam)
```

## Step 4: Effect Size (η²)

```{r}
ss_exam  <- summary(model_exam)[[1]][["Sum Sq"]]
eta_exam <- ss_exam[1] / sum(ss_exam)
cat("Eta-squared =", round(eta_exam, 3), "\n")
cat("Benchmarks: small = 0.01, medium = 0.06, large = 0.14\n")
```

## Step 5: Post-Hoc (Tukey HSD)

```{r}
TukeyHSD(model_exam)
```

## Step 6: Plot

```{r fig.width=6, fig.height=5}
boxplot(score_exam ~ school, data = df_exam,
        col  = c("lightyellow", "lightblue", "lightgreen"),
        main = "Exam Performance by School Type",
        xlab = "School Type", ylab = "Exam Score", las = 1)

means_exam <- tapply(df_exam$score_exam, df_exam$school, mean)
points(1:3, means_exam, pch = 18, col = "red", cex = 1.5)
legend("bottomright", legend = "Group mean", pch = 18, col = "red")
```

---

# Summary

| Question | Design | Normality | Variance Check | Test | Post-Hoc |
|---|---|---|---|---|---|
| Q1 Memory | Independent | `shapiro.test` | `bartlett.test` / `levene_manual` | `aov` | `TukeyHSD` |
| Q2 Cond.1 Driving | Repeated Measures | `shapiro.test` | `mauchly.test` (sphericity) | `aov` + `Error()` | `pairwise.t.test` (Bonferroni) |
| Q2 Cond.2 Driving | Independent | `shapiro.test` | `bartlett.test` / `levene_manual` | `aov` | `TukeyHSD` |
| Q3 Exam | Independent | `shapiro.test` | `bartlett.test` / `levene_manual` | `aov` | `TukeyHSD` |

**All code uses base R only — no extra packages required.**
