---
title: "Anscombe_rmd_code"
output:
  pdf_document: default
  html_document: default
---

Name: Rachakonda Hrithik Sagar 


Roll No: 202390021

```{r}
data <- read.csv("Anscombe_dataset.csv")
str(data)
```


#descriptive stats
```{r}
library(dplyr)

data %>%
  group_by(Group) %>%
  summarise(
    mean_x = mean(x),
    mean_y = mean(y),
    sd_x = sd(x),
    sd_y = sd(y),
    cor_xy = cor(x, y)
  )
```
#plots
```{r}
par(mfrow = c(2,2))
groups <- unique(data$Group)

for(g in groups){
  subset_data <- subset(data, Group == g)
  plot(subset_data$x, subset_data$y,
       main = paste("Group", g),
       xlab = "X", ylab = "Y",
       pch = 19)
  abline(lm(y ~ x, subset_data), col = "red")
}


```



Observations

	•	Overall, the mean, standard deviation, and correlation are approximately the same across all four groups. However, the underlying data distributions and relationships differ substantially.

	•	Group 1: A clear linear relationship is observed between the variables X and Y.

	•	Group 2: A relationship between X and Y exists, but it is non-linear, which is not captured well by correlation alone.

	•	Group 3: A generally linear relationship is present, but it is strongly influenced by a single outlier, which disproportionately affects the correlation.

	•	Group 4: There is no meaningful relationship between X and Y; the mean, standard deviation, and correlation coefficient are largely driven by a single outlier.

